---
title: "Glycan Reachability Analysis with GTEx Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Glycan Reachability Analysis with GTEx Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Overview

This vignette demonstrates how to use `glycoreach` to compute glycan
biosynthetic reachability scores from the GTEx v8 gene expression dataset.

## Installation

```{r install}
# install.packages("remotes")
remotes::install_github("matsui-lab/glycoreach")
```

## Load data

Download the GTEx v8 gene median TPM file from the
[GTEx Portal](https://gtexportal.org/home/downloads/adult-gtex/bulk_tissue_expression):

```{r load}
library(glycoreach)
library(data.table)

# Read GTEx gene-level TPM (GCT format, skip first 2 header lines)
gct <- fread("GTEx_Analysis_2017-06-05_v8_RNASeQCv1.1.9_gene_median_tpm.gct.gz",
             skip = 2)

# Extract gene symbols and TPM matrix
gene_symbols <- gct$Description
tpm_cols <- setdiff(names(gct), c("Name", "Description"))
tpm <- as.matrix(gct[, ..tpm_cols])
rownames(tpm) <- gene_symbols

# Handle duplicate gene symbols (average)
if (any(duplicated(rownames(tpm)))) {
  tpm <- do.call(rbind, lapply(split(seq_len(nrow(tpm)), rownames(tpm)),
                               function(idx) colMeans(tpm[idx, , drop = FALSE])))
}

dim(tpm)
# [1] ~56200 genes x 54 tissues
```

## Compute reachability scores

```{r compute}
# Compute all 22 reachability metrics
res <- compute_reachability(tpm)
print(res)

# Compute only sLeX
res_slex <- compute_reachability(tpm, pathways = "slex")
```

## Visualize tissue variation

```{r viz}
library(ggplot2)

# sLeX reachability across tissues
plot_data <- res[, .(tissue = sample, sLeX = sLeX_reachability)]
plot_data <- plot_data[order(sLeX)]
plot_data[, tissue := factor(tissue, levels = tissue)]

ggplot(plot_data, aes(x = tissue, y = sLeX)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(title = "sLeX Reachability across GTEx Tissues",
       x = NULL, y = "Reachability Z-score") +
  theme_minimal()
```

## Knockout simulation

Simulate the effect of knocking out FUT8 (core fucosyltransferase)
on N-glycan core fucosylation:

```{r ko}
# Wild-type
res_wt <- compute_reachability(tpm, pathways = "nglycan")

# FUT8 knockout
res_ko <- compute_reachability(tpm, pathways = "nglycan", ko_genes = "FUT8")

# Compare core fucosylation scores
comparison <- data.table(
  tissue = res_wt$sample,
  WT = res_wt$Ng_coreFuc,
  KO = res_ko$Ng_coreFuc,
  delta = res_ko$Ng_coreFuc - res_wt$Ng_coreFuc
)
print(comparison[order(delta)])
```

## Using pre-computed reference Z-scores

When comparing new samples against the GTEx reference, use GTEx-derived
mean and SD for normalization:

```{r reference}
# Compute reference statistics from GTEx (full sample-level data)
full_tpm <- fread("GTEx_Analysis_2017-06-05_v8_RNASeQCv1.1.9_gene_tpm.gct.gz",
                  skip = 2)
# ... extract matrix as above ...

L_ref <- log1p(full_tpm_matrix)
ref_mu <- rowMeans(L_ref, na.rm = TRUE)
ref_sd <- apply(L_ref, 1, sd, na.rm = TRUE)
ref_sd[ref_sd == 0] <- 1

# Apply to new data using GTEx reference
new_res <- compute_reachability(new_tpm, mu = ref_mu, sd = ref_sd)
```

## Pathway definitions

View the built-in pathway definitions:

```{r defs}
defs <- load_pathway_definitions()
names(defs$pathways)
# [1] "sLeX" "gangliosides" "heparan_sulfate" "N_glycan_processing" "O_GalNAc"

# Inspect sLeX pathway
str(defs$pathways$sLeX)
```
