---
title: "Single-Cell Reachability with Tabula Sapiens"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Single-Cell Reachability with Tabula Sapiens}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Overview

This vignette demonstrates how to apply `glycoreach` to single-cell RNA-seq
data from the [Tabula Sapiens](https://tabula-sapiens-portal.ds.czbiohub.org/)
human cell atlas. The key strategy is **pseudobulk aggregation**: averaging
expression within each cell type to mitigate dropout noise while preserving
cell-type-specific signatures.

## Prerequisites

```{r deps}
# install.packages("remotes")
remotes::install_github("matsui-lab/glycoreach")

# For reading h5ad files
if (!requireNamespace("anndata", quietly = TRUE))
  reticulate::py_install("anndata")
if (!requireNamespace("zellkonverter", quietly = TRUE))
  BiocManager::install("zellkonverter")
```

## Load Tabula Sapiens data

Download the Tabula Sapiens h5ad file from
[CZ CELLxGENE](https://cellxgene.cziscience.com/collections/e5f58829-1a66-40b5-a624-9046778e74f5):

```{r load}
library(glycoreach)
library(data.table)
library(zellkonverter)
library(SingleCellExperiment)

# Read the h5ad file
sce <- readH5AD("TabulaSapiens.h5ad")

# Check available metadata
head(colData(sce))
# Key columns: organ_tissue, cell_ontology_class, donor
```

## Pseudobulk aggregation

Aggregate counts by cell type within each organ to create robust
expression profiles:

```{r pseudobulk}
# Define grouping: organ + cell type
groups <- paste(sce$organ_tissue, sce$cell_ontology_class, sep = "|")

# Get raw counts and compute pseudobulk TPM
counts <- counts(sce)
gene_lengths <- rowData(sce)$feature_length  # if available

# Simple pseudobulk: sum counts per group, then convert to TPM-like
group_levels <- unique(groups)
pseudobulk <- matrix(0, nrow = nrow(counts), ncol = length(group_levels),
                     dimnames = list(rownames(counts), group_levels))

for (gl in group_levels) {
  idx <- which(groups == gl)
  if (length(idx) < 10) next  # skip groups with < 10 cells
  pseudobulk[, gl] <- rowSums(counts[, idx])
}

# Remove empty columns
pseudobulk <- pseudobulk[, colSums(pseudobulk) > 0]

# Normalize to CPM (counts per million) as TPM proxy
cpm <- t(t(pseudobulk) / colSums(pseudobulk) * 1e6)
```

## Compute reachability

```{r compute}
# Compute reachability across all cell types
res <- compute_reachability(cpm)

# Parse organ and cell type from sample names
res[, c("organ", "cell_type") := tstrsplit(sample, "\\|", fixed = FALSE)]

print(res[, .(organ, cell_type, sLeX_reachability, GM3_reach, reach_FGF_like)])
```

## Visualize cell-type variation

```{r viz_organ, fig.width=10, fig.height=8}
library(ggplot2)

# sLeX reachability by cell type within selected organs
organs_of_interest <- c("Lung", "Blood", "Liver", "Pancreas")
plot_data <- res[organ %in% organs_of_interest]
plot_data <- plot_data[order(sLeX_reachability)]
plot_data[, label := paste0(organ, ": ", cell_type)]
plot_data[, label := factor(label, levels = label)]

ggplot(plot_data, aes(x = label, y = sLeX_reachability, fill = organ)) +
  geom_col() +
  coord_flip() +
  labs(title = "sLeX Reachability by Cell Type (Tabula Sapiens)",
       x = NULL, y = "Reachability Z-score") +
  theme_minimal() +
  theme(legend.position = "bottom")
```

## Compare cell types across organs

```{r heatmap, fig.width=12, fig.height=8}
# Heatmap of all metrics by cell type
metrics <- c("sLeX_reachability", "GM3_reach", "GM1_reach", "GD3_reach",
             "reach_FGF_like", "reach_WNT_like",
             "Ng_complex", "Ng_sia", "OGN_Core1")

mat <- as.matrix(res[, ..metrics])
rownames(mat) <- res[, paste0(organ, "|", cell_type)]

# Use pheatmap or ComplexHeatmap for visualization
if (requireNamespace("pheatmap", quietly = TRUE)) {
  pheatmap::pheatmap(mat, scale = "column",
                     clustering_distance_rows = "euclidean",
                     main = "Glycan Reachability across Cell Types",
                     fontsize_row = 6)
}
```

## Comparison: bulk GTEx vs single-cell resolution

A key advantage of single-cell reachability is resolving heterogeneity
hidden in bulk data. For example, bulk liver has moderate sLeX reachability,
but at single-cell resolution this decomposes into:

- **Hepatocytes**: low sLeX (limited FUT3/5/6 expression)
- **Kupffer cells**: high sLeX (immune-lineage expression pattern)
- **Endothelial cells**: moderate sLeX

This heterogeneity is invisible in bulk GTEx data but directly accessible
via pseudobulk reachability.

## Extending to Tabula Muris Senis (aging)

The same workflow applies to the mouse aging atlas
[Tabula Muris Senis](https://tabula-muris-senis.ds.czbiohub.org/).
Use ortholog mapping to convert mouse gene symbols to human:

```{r mouse}
# After loading Tabula Muris Senis data and creating pseudobulk...
# Map mouse symbols to human orthologs

# Using biomaRt (example)
if (requireNamespace("biomaRt", quietly = TRUE)) {
  human <- biomaRt::useEnsembl("genes", dataset = "hsapiens_gene_ensembl")
  mouse <- biomaRt::useEnsembl("genes", dataset = "mmusculus_gene_ensembl")

  orthologs <- biomaRt::getLDS(
    attributes = "mgi_symbol", filters = "mgi_symbol",
    values = rownames(mouse_cpm), mart = mouse,
    attributesL = "hgnc_symbol", martL = human
  )

  # Rename rows to human symbols and compute reachability
  # (handle many-to-many mappings by averaging)
}
```

## Notes

- **Minimum cell count**: We recommend at least 10 cells per pseudobulk
  group. Groups with fewer cells are unreliable due to dropout noise.
- **Normalization**: CPM from pseudobulk counts is a reasonable TPM proxy
  for reachability analysis. If gene lengths are available, use proper TPM.
- **Z-score reference**: For cross-dataset comparisons, use GTEx-derived
  mean/SD as reference (see GTEx vignette).
